---
title: "Airline Answers"
format: html
---

1.  Create a new data set, `flights2` that contains the carrier name, year, month, day, departure delay, arrival delay, origin airport, destination airport, and flight number.

```{r}
flights2 <- flights %>%
  select(carrier, year, month, day, dep_delay, arr_delay, origin, dest, flight)
```

2.  Filter the data set to only show United Airlines flights. What month in 2013 did United Airlines have the most flights from the New York Area?

```{r}
flightsUA <- flights2 %>% filter(carrier == "UA")

flightsUA %>%
  group_by(month) %>%
  summarize(N= n()) %>%
  arrange(desc(N))

```

3.  How many unique destinations does United Airlines serve from the New York Area?

```{r}
flightsUA %>%
  group_by(dest) %>%
  summarise(N = n())
```

4.  How many unique unique destinations does United Airlines run from each of the three area airports?

```{r}
flightsUA %>%
  group_by(origin, dest) %>%
  summarise(N = n())


flightsUA %>%
  group_by(origin) %>%
  summarise(N = length(unique(dest)))
```

5.  What is the average departure delay of a United Airlines flight leaving any New York area airport?

```{r}
flightsUA %>%
  filter(!is.na(dep_delay)) %>%
  summarize(dep_delay = mean(dep_delay))
```

6.  What is the average departure delay of a United Airlines flight leaving JFK? LGA? EWR?

```{r}
flightsUA %>%
  filter(!is.na(dep_delay)) %>%
  group_by(origin) %>%
  summarize(dep_delay = mean(dep_delay))
```
