---
title: 'Section 6: Data Classes and tidyr'
author: 'MTH 365: Introduction to Data Science'
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r}
library(tidyverse)
```



# More with Data Cleaning: `tidyr()`

Remember: “Tidy” data is a standard way of mapping the meaning of a data set to its structure.

1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table.

Any other arrangement of the data is called “messy”.

Real data sets can, and often do, violate the three principles of tidy data in almost every way imaginable! 

### What makes Data `tidy`? Key-Value Pairs

```{r, echo = FALSE}
example <- data.frame(Patient = c(rep("John Smith",2), rep("Jane Doe",2), rep("Mary Johnson",2)), Treatment = rep(c("A", "B"), 3), Score = c(NA, 18, 4, 1, 6, 7))

example
```

- Treatment and Patient uniquely describe a single row in the data set
- Treatment and Patient are key variables
- Score is a measurement variable
- This makes Treatment-Patient and Score a key-value pair

Key-Value pairs (KVP) - also attribute-value, field-value, name-value: abstract data representation that allows a lot of flexibility

One way of telling whether a data set is tidy is to check that all keys for a value are aligned in one row.

### Example: Untidy Data

```{r}
#install.packages("reshape2") - if don't have installed already
library(reshape2)
data("french_fries")
head(french_fries)

```


This format is not ideal for Data Analysis

```{r}
ggplot(french_fries) + 
  geom_boxplot(aes(x="1_buttery", y=buttery), fill = "cyan4") +
  geom_boxplot(aes(x = "2_grassy", y = grassy), fill = "darkorange2") +
  geom_boxplot(aes(x = "3_painty", y = painty), fill = "darkorchid1") +
  geom_boxplot(aes(x = "4_potato", y = potato), fill = "chartreuse3") +
  geom_boxplot(aes(x = "5_rancid", y = rancid), fill = "deeppink") +
  xlab("variable") + ylab("rating")
```


### Tidy your data using `pivot_longer`

When pivoting longer, you need to specify:

- the `cols` (identifiers)
- the `names_to` (new column name)
- the `values_to` (measures from cols)

```{r}
french_fries_long <- french_fries %>% 
  pivot_longer(cols = potato:painty, #5:9
               names_to = "variable", 
               values_to = "rating")

head(french_fries_long)
```

This format is better for Data Analysis

```{r}
ggplot(french_fries_long) + 
  geom_boxplot(aes(x = variable, y = rating, fill = variable))
```


### And reverse: `pivot_wider`

Useful if we want to display things as a table

When pivoting wider, you need to specify:

- the `names_from` (column name that we want to widen)
- the `values_from` (measures from column we want to widen)

```{r}
french_fries_wide <- french_fries_long %>% 
  pivot_wider(names_from = variable, 
              values_from = rating, 
              values_fill = NA)
head(french_fries_wide)
```

Now we are back to the original data set

# Data types in R

A common type of error is caused by the data type. This is tricky because it may not directly give you the error message!

R is (usually) good at figuring out the best data type, but sometimes we'll need to take corrective action! The five data types we'll interact with the most are:

- Logical
- Double
- Integer
- Character
- Lists

### Logical 

```{r}

```

### Character 

```{r}

```

### Double 

```{r}

```

### Integer

```{r}

```

### Lists

```{r}

```


### Named lists

Because of their more complex structure we often want to name the elements of a list (we can also do this with vectors). This can make reading and accessing the list more straight forward.

```{r}

```

### Concatenation

Vectors can be constructed using the `c()` function.

```{r}

```


### Coercion

When you mix data types within a vector, R will create the result with the data type that can most easily accommodate all the elements it contains. This conversion between modes of storage is called “coercion”. 

- When R converts the mode of storage based on its content, it is referred to as “implicit coercion”. Can you guess the following data types (without running them first)?

```{r}

```

Use the `typeof()` function to find out.

```{r}

```


### Missing values

R uses `NA` to represent missing values in its data structures.

```{r}

```

### Other special values

- `NaN`: Not a number
- `Inf`: Positive infinity
- `-Inf`: Negative infinity

```{r}

```


### Example: Cat lovers

A survey asked respondents their name and number of cats. The instructions said to enter the number of cats as a numerical value.

```{r message=FALSE, echo=FALSE}
# Make sure to change the file directory
cat_lovers <- read.csv("../Week 6/data/cat_lovers.csv")
glimpse(cat_lovers)
```

Let's try to see the average number of cats.

```{r}
cat_lovers %>%
  summarise(mean_cat = mean(number_of_cats))
```

How about removing the NA value?

```{r}
cat_lovers %>%
  summarise(mean_cat = mean(number_of_cats, na.rm = TRUE))
```

Slow down.

What is the type of the `number_of_cats` variable?
  
```{r}
typeof(cat_lovers$number_of_cats)
```

- Let's look at the actual data, are there any other strange responses?

```{r}
cat_lovers
```

If the data is too large to look at all lines. Since you have already know what is the problem,

```{r}
tail(cat_lovers, n = 10)
```

Problem: `number_of_cats` is a characters, when we want it to be a number. However, we have some strange response.

**Attempt 1**: convert to numeric, where the respones that couldn't be converted to a number become NA

```{r}
cat_lovers1 <- cat_lovers %>% 
  mutate(number_of_cats_new = as.numeric(number_of_cats))

str(cat_lovers1)

cat_lovers1 %>% filter(is.na(number_of_cats_new))

cat_lovers1 %>% filter(is.na(number_of_cats_new)) %>%
  summarize(mean(number_of_cats_new, na.rm = TRUE))
```

You can discard these two lines, but they are still useful information, just in different format. This is why surveys restrict entry values...How about we convert them with `case_when()` function. 

**Attempt 2** : define a new variable?

```{r, eval = FALSE}
cat_lovers %>%
  mutate(
    number_of_cats = case_when(
      name == "Ginger Clark" ~ 2,
      name == "Doug Bass"    ~ 3
    ),
  ) %>%
  summarise(mean_cats = mean(number_of_cats))
```

How about this one, what is the problem?


```{r}
cat_lovers %>%
  mutate(
    number_of_cats = case_when(
      name == "Ginger Clark" ~ 2,
      name == "Doug Bass"    ~ 3
    )
  ) %>% tail(n=10)
```

Now we have our data in two separate columns

Now that we (maybe) know what we're doing...

```{r, eval = FALSE}
cat_lovers2 <- cat_lovers %>%
  mutate(number_of_cats = case_when(
    name == "Ginger Clark" ~ 2,
    name == "Doug Bass"    ~ 3,
    .default = as.numeric(number_of_cats)
    )
  )

cat_lovers2 %>% tail(n=10)

cat_lovers2 %>% summarise(mean_cats = mean(number_of_cats))
```


You may want to save a copy of this new cleaned version of data

```{r, eval = FALSE}

glimpse(cat_lovers2)

write.csv(cat_lovers2, file = "cat_lovers_clean.csv", row.names = FALSE)
```

### Moral of the story

- If your data doesn't behave how you expect it to, implicit coercion a might be the reason.
- Go in and investigate your data, apply the fix, _save your data_, live happily ever after.


## Data "sets" in R

- "set" is in quotation marks because it is not a formal data class

A tidy data "set" can be one of the following types:

- `tibble`
- `data.frame`

We'll often work with `tibble`s:

- `readr` package (e.g. `read_csv` function) loads data as a `tibble` by default
- `tibble`s are part of the tidyverse, so they work well with other packages we are using
- They make minimal assumptions about your data, so are less likely to cause hard to track bugs in your code


### Data frames

- "set" is in quotation marks because it is not a formal data class

A tidy data "set" can be one of the following types:

- `tibble`
- `data.frame`

We'll often work with tibbles:

- `readr` package (e.g. `read_csv` function) loads data as a tibble by default
- `tibbles` are part of `tidyverse`, so they work well with other packages we are using
- They make minimal assumptions about your data, so are less likely to cause hard to track bugs in your code

A data frame is the most commonly used data structure in R, they are just a list of equal length vectors (usually atomic, but you can use generic as well). 

- Each vector is treated as a column and elements of the vectors as rows.

A `tibble` is a type of data frame that ... makes your life (i.e. data analysis) easier.

- Most often a data frame will be constructed by reading in from a file, but we can also create them from scratch.


### Example: How many respondents have below average number of cats?

```{r}
mean_cats <- cat_lovers %>%
  summarize(mean_cats = mean(number_of_cats))

cat_lovers %>% 
  filter(number_of_cats < mean_cats) %>%
  nrow()
```

Have you seen any problem here?

```{r}
mean_cats
class(mean_cats)
```

**A possible solution**: `pull()`.

`pull()` works like [[]] for data frames, and pulls out the value of a single variable in a data frame. How does `pull()` work?

```{r}
cat_lovers %>% pull()
#blank pulls last?
cat_lovers  %>% pull(2)
cat_lovers  %>% pull(number_of_cats)
```

`pull()` can be your new best friend, when used correctly.

```{r}
mean_cats <- cat_lovers %>%
  summarize(mean_cats = mean(number_of_cats)) %>%
  pull()

mean_cats
typeof(mean_cats)

cat_lovers %>% 
  filter(number_of_cats < mean_cats) %>%
  nrow()
```

## Factors

Factor objects are how R stores data for categorical variables (fixed numbers of discrete values).

```{r}
x = factor(c("BS", "MS", "PhD", "MS"))
x
```

```{r}
glimpse(x)
```

```{r}
class(x)
```

```{r}
glimpse(cat_lovers)

ggplot(cat_lovers,aes(x = handedness)) +
  geom_bar()
```

What if we want to force a particular order?

```{r}
cat_lovers <- cat_lovers %>%
  mutate(handedness = fct_relevel(handedness, "left", "ambidextrous", "right"))
glimpse(cat_lovers$handedness)

ggplot(cat_lovers,aes(x = handedness)) +
  geom_bar()
```

## Factor notes

- R uses factors to handle categorical variables, variables that have a fixed and known set of possible values. Historically, factors were much easier to work with than character vectors, so many base R functions automatically convert character vectors to factors.

- However, factors are still useful when you have true categorical data, and when you want to override the ordering of character vectors to improve display. The goal of the forcats package is to provide a suite of useful tools that solve common problems with factors.

Source: [forcats.tidyverse.org](http://forcats.tidyverse.org/)

### Overriding Data Types


If you are absolutely sure of a data class, overwrite it in your tibble so that you don't need to keep having to keep track of it

```{r}
x <- "2"
typeof(x)
```

```{r}
x <- as.numeric(x)
typeof(x)

```

### Overriding Data Types: Weird Things About `R`

```{r}
cat_lovers$handedness <- as.factor(cat_lovers$handedness)
class(cat_lovers$handedness)
typeof(cat_lovers$handedness)
```

The `typeof function` is giving information that's at a "lower" level of abstraction. Factor variables (and also Dates) are stored as integers. 

  - Determines the (R internal) type or storage mode of any object
  
- `class`: a simple generic function mechanism which can be used for an object-oriented style of programming. 

So when working with data sets use `str`, `glimpse`, `class`

## Recap

Always best to think of data as part of a tibble

  - This plays nicely with the `tidyverse` as well
  - Rows are observations, columns are variables

Be careful about data types/classes

  - Sometimes `R` makes silly assumptions about your data class 
  - Using `tibble`s help, but it might not solve all issues
  - Think about your data in context, e.g. 0/1 variable is most likely a `factor`
  - If a plot/output is not behaving the way you expect, first investigate the data class
  - If you are absolutely sure of a data class, overwrite it in your tibble so that you don't need to keep having to keep track of it
  - `mutate` the variable with the correct class
  
  
  
# R Debugging Tips

Like all the other programming language, R sometimes will generate errors or the output may not be the one you want. There are multiple reasons for that. In this course, we don't expect you to write functions so in general, that makes the debugging easier. 

Here is some common type of errors: 

1. "could not find function" errors, usually caused by typos or not loading a required package

```{r, eval=FALSE}
glimpse(iris)
```

2. "Error in if" errors, caused by non-logical data or missing values passed to R's "if" conditional statement

```{r, eval = FALSE}
if(c(1,2,3) > 2) {
  print("True")
}
```

3. "Error in eval" or "object not found" errors, caused by references to objects that don't exist

```{r, eval = FALSE}
abc
```

4. "cannot open" errors, caused by attempts to read a file that doesn't exist or can't be accessed

```{r, eval = FALSE}
cat_lover = read.csv("cat_lovers.csv")
```

4. "no applicable method" errors, caused by using an object-oriented function on a data type it doesn't support

```{r, eval = FALSE}
library(tidyverse)
head(cat_lover)
cat_lover %>% summarise(mean(handedness))
```

5. "subscript out of bounds" errors, caused by trying to access an element or dimension that doesn't exist

```{r, eval = FALSE}
cat_lover %>% summarise(mean(handness))
```

6. package errors caused by being unable to install, compile or load a package.

```{r, eval = FALSE}
library(errorist)
```

Here is a few strategies if your code does not work:

1. Google!

Whenever you see an error message, start by googling it. If you’re lucky, you’ll discover that it’s a common error with a known solution. When googling, improve your chances of a good match by removing any variable names or values that are specific to your problem.

2. Make sure the problem does exist.

Sometimes the error may not cause by the current lines you are working on. It is possible that you have changed something earlier. Try to re-run the files from the beginning.

3. Run the functions part by part till you find the problem. 

This is especially useful for the tidyverse functions. 

```{r, eval = FALSE}
cat_lover %>% filter(handedness == "left") %>% 
  ggplot(aes(x = number_of_cats)) + 
  geom_histogram()

glimpse(cat_lover)
```

4. Print out the output to see whether it is the one you want. Sometimes look at the data type may also help you understand what's going on. 
